Designing and modeling a browser-based distributed evolutionary computation systems

### 2015-04-08 modeling-volunteer-computing-gecco.tex
Improving client-side description

% Rewrite this - JJ
In order to test the volunteer computing environment, a presentation
describing a low cost volunteer computing environment was
created. This presentation was actually delivered in several
conferences (such as IDC 2014). During the %FERGU: TODO JJ, no te
                                %olvides de decirme esto. Es la IDC
                                %2014?
% Y el PPSN. Pero eso lo voy a dejar para el trabajo anterior, voy a
% intentar hacer experimentos nuevos. - JJ
conference, it was revealed to the persons attending it that they were participating
in the experiment. The same procedure was used when trying to gather
users in social networks: a description of the container (the
presentation) and disclosure of its purpose. The upper right corner of
the presentation shows the progress of the evolution. It
stops when the solution has been reached.

Above the experiment, from the point of view of the user, has been
described. What it is actually running is an island model
\cite{muhlenbein1991parallel} in every browser that uses the server as
a {\em shuttle} to transfer individuals from one browser to another in
what can be eventually a fully connected topology; however, all
connections take place between the browser and the server.
This deals with
several problems. The connection is stateless: islands connect to the
server to send and receive a single individual but there is no task
balancing: all islands start clicking on an URL and finish when they
browse off to the next page. Fault tolerance is implicit, in the sense
that no island has information that cannot be, or generated, somewhere
else, although if the server fails the experiment results might be
lost. All operation is asynchronous, with islands entering and leaving
the experiment all of their own.

We will next explain its different parts from the point of
view of the architecture itself and the algorithm that it is actually
running. First, we will describe the client code and next, in
subsection \ref{ss:server} the server architecture and where it is
hosted.

@@

As stated above, the evolutionary algorithms run mainly in the
browser. The problem run is a multimodal problem called {\em l-trap},
which has been used extensively as a benchmark for evolutionary
algorithms \cite{fernandes2009using,nijssen2003analysis}. This
function counts the number of bits in a sequence of $l$ and assigns
the local maximum if it has got 0 bits and the global maximum to $l$
bits. The fitness goes down to a {\em trap} as you increase the number
of bits, that is why it creates a rugged fitness landscape that is
difficult to surmount for evolutionary algorithms. Its difficulty is
increased as the number of traps concatenated grows, so that in some
cases it might need millions of evaluations to find the solution. In
our case we have used several values, from 30 to 50 traps, although we
were not so much focused on finding the solution (or the fastest way
of finding it) but on creating a experiment that could last for a
certain amount of time, around one hour, so that there would be a
chance for having many users checkout out the page and contributing
cycles to it.

This fitness is implemented as part of the Nodeo evolutionary
algorithm library written in node.js, a server-based implementation of
JavaScript. However, since it is going to be run on the browser, a
procedure called ``browserification'' must be applied to it so that it
can run, without modifications, in any browser. The main difference
between JavaScript on the browser and node.js is the way they load the
external libraries. The algorithm used a population of 256
individuals, random initialization, a canonical evolutionary algorithm
with an elite of 2 individuals and crossover and then mutation applied
to all the population pool, that was generated using 2-individual
tournament selection. Once again, the particular evolutionary
algorithm used is not so important, since we were rather focused on
the user behavior, although it is quite clear that the selective
pressure of the algorithm will have an impact in its performance and
will play for or against its asynchrony in many different and complex
ways \cite{jj:2008:PPSN}.

The initial code was only modified to represent the fitness in a chart
(that uses the {\tt chart.js} JavaScript library) so that the user has
a visual feedback of what is going on. The {\em migration} part is
also added: after every 100 generations, a single individual is sent
to the server and another request is made to the server of a single
and random individual. This makes the procedure less greedy, and also
decreases the probability of getting the same individual that was sent
some generations back. These two operations are made in {\em fire and
  forget} style, that is, without any error checking, so that if the server is
down, or indeed if the algorithm is run locally, it is not interrupted
and continues operation until the solution is found, a string with all
ones.

As a summary, the only thing needed to run an evolutionary algorithm
in the browser is to add {\em migration} functions. To make the
operation of the algorithm independent of them is a plus, because it
makes the local algorithm more fault-tolerant. The algorithm running
on every browser has the same parameters and starts from a random
population when the page is loaded.

>>>>

% Rewrite this - JJ
In order to test the volunteer computing environment, a presentation
describing a low cost volunteer computing environment was
created. This presentation was actually delivered in several
conferences (such as IDC 2014, PPSN 2014 and FOSDEM 2015). These
results have been presented in \cite{arxiv:modeling}. In this paper,
although it uses exactly the same evolutionary algorithm client code,
the {\em presentation} has changed: it uses a single page with several
panels showing the progress of the algorithm and how many users have
been contributing to the experiment. To get users, the URL
(\url{http://nodio-jmerelo.rhcloud.com} and
\url{http://nodeo-i.herokuapp.com}) where presented, with a short
message, in Twitter and other social networks, introducing it as a
``demo for evolutionary algorithms'', sometimes in Spanish, some in
English.

Besides visualizing the evolution, the browser is actually running an
distributed evolutionary algorithm that follows an  island model
\cite{muhlenbein1991parallel} using the server to transfer individuals from one browser to another in
what can be eventually a fully connected topology, although the term
does not correctly apply here.

We will next explain the two framework tiers from the point of
view of the architecture itself and the algorithm that it is actually
running. First, we will describe the client code and next, in
subsection \ref{ss:server} the server architecture and where it is
hosted.

@@

As stated above, the evolutionary algorithms run mainly in the
browser. The problem run is a multimodal problem called {\em l-trap},
which has been used extensively as a benchmark for evolutionary
algorithms \cite{fernandes2009using,nijssen2003analysis}. This
function counts the number of bits in a sequence of $l$ and assigns
the local maximum if it has got 0 bits and the global maximum to $l$
bits. The fitness falls into a {\em trap} as you increase the number
of bits, that is why it creates a rugged fitness landscape that is
difficult to surmount for evolutionary algorithms. Its difficulty is
increased as the number of  concatenated traps grows, turning it into
a difficult problem for evolutioanry algorithms. In
our case we have used $l=40$ traps.

This fitness is implemented as part of the NodEO evolutionary
algorithm library written in node.js \cite{DBLP:conf/gecco/GuervosVGES14}, a server-based implementation of
JavaScript, converted to a client-based version via
``browserification''. The algorithm runs a canonican evolutionary
algorithm on a population of 128, which is in fact too small for a
single node to find the solution; at least a population of 512 has
been found to be necessary to find the solution with a certain
probability. Initialization is random initialization and a canonical evolutionary algorithm
with an elite of 2 individuals is run, with crossover and then mutation applied
to all the population pool generated using 2-individual
tournament selection. The particular evolutionary
algorithm used is not so important, since we were rather focused on
the user behavior, although it is quite clear that the selective
pressure of the algorithm will have an impact in its performance and
will play for or against its asynchrony in many different and complex
ways \cite{jj:2008:PPSN}.

The initial code was only modified to represent the fitness in a chart
(that uses the {\tt chart.js} JavaScript library) so that the user has
a visual feedback of what is going on. The {\em migration} part is
also added: after every 100 generations, a single individual is sent
to the server and another request is made to the server of a single
and random individual; this algorithm can be also run locally, the
algorithm does not stop if there is no answer from the server. As a
summary, the only thing needed to run an evolutionary algorithm in the browser is to add {\em migration} functions to the original one.
